---
title: "HW 10"
author: "Lily Walker"
format: pdf
editor: visual
---

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(rvest)
```

## Question 1

Provide R code to create a data frame containing the data from the States by Size in Square Miles table from the following URL: https://statesymbolsusa.org/symbol-official-item/national-us/uncategorized/states-size Make sure to properly read in the column names in R. Also, omit the unnecessary index column in the state size data set. Plot a histogram of the state sizes.

```{r}
#| message: false
#| warning: false
url1 <- 'https://statesymbolsusa.org/symbol-official-item/national-us/uncategorized/states-size'

states <- read_html(url1) |> 
  html_element('table') |> 
  html_table(header = T) |> 
  setNames(c('x', 'state', 'square_miles')) |> 
  select(state, square_miles)

states
```

## Question 2

Find a website that contains a “table” html element. Get the URL and scrape the table into a dataframe using the methods we covered in class. Make a well-labeled graphic based on the data.

```{r}
url2 <- 'https://cotap.org/per-capita-carbon-co2-emissions-by-country/?gad_source=1&gad_campaignid=15416182513&gbraid=0AAAAADyLLdeJKbU3OAvndOx5THL3tU_ty&gclid=Cj0KCQjwp7jOBhDGARIsABe7C4eNFMW0j86sKV6qYsMvpJwbCzBJHt0SeIy2ryjPratPNd_7U265U9AaAosqEALw_wcB'

co2 <- read_html(url2) |> 
  html_element('table') |> 
  html_table() |> 
  setNames(c('country', 'co2_per_cap')) |> 
  mutate(co2_per_cap = as.numeric(co2_per_cap))

english_countries <- c('United States', 'Australia', 'United Kingdom', 'Canada', 'Ireland', 'New Zealand')

co2 |> 
  filter(country %in% english_countries) |> 
  ggplot(aes(x = country, y = co2_per_cap, fill = country)) +
  geom_col(show.legend = F) +
  geom_hline(
    yintercept = mean(co2$co2_per_cap, na.rm = TRUE),
    color = "red",
    linewidth = 1.5
    ) +
  labs(
    title = 'CO2 Emissions Per Capita',
    subtitle = 'Among native english speaking countries',
    x = 'Country',
    y = 'Per Capita Carbon Emissions(tons)',
    caption = '* red line = world average'
  )
```

## Question 3

The following website contains a table with information about the spiciness of peppers: https: //trueleafmarket.com/pages/scoville-heat-scale?srsltid=AfmBOoot8CjsXlDYBFF6aAwbqC AaHg5qEPpsbdOn9U8wzZIwwiL0MxZ9 Create the following plot using these data

```{r fig.width=8, fig.height=8}
url3 <- 'https://trueleafmarket.com/pages/scoville-heat-scale?srsltid=AfmBOoot8CjsXlDYBFF6aAwbqCAaHg5qEPpsbdOn9U8wzZIwwiL0MxZ9'

peppers <- read_html(url3) |> 
  html_element('table') |> 
  html_table() |> 
  janitor::clean_names() |> 
  mutate(min_SHU = parse_number(average_scoville_heat_units))

category_inds <- which(str_sub(peppers$average_scoville_heat_units, 1, 1) != "≈")
categories <- peppers$pepper_variety[category_inds]

peppers_clean <- peppers[-category_inds,] |> 
  mutate(Category = rep(categories, 
                        diff(c(category_inds, nrow(peppers) + 1)) - 1))

peppers_clean |> 
  filter(Category %in% c('Very Hot', 'Hottest Peppers on Earth', 
                         'Even Spicier Chemicals')) |> 
  ggplot(aes(y = fct_reorder(pepper_variety, min_SHU), x = min_SHU)) +
  geom_col(aes(fill = Category)) +
  labs(x = "Minimum Scoville Heat Units",
       y = "Pepper Variety") 
```
